A spiral galaxy's mass distribution uncovered through lensing and dynamics

### 2016-06-08 LensingPaper_draft_v4.01.tex
Wrote section that speculates on the matter distribution of J1331.

%--------------------------------------------
\section{Introduction}
%--------------------------------------------

@@

%--------------------------------------------
\Wilma{[Referee says: "The manuscript in its current form is also slightly longer than necessary at certain
points: [...] Section 3.3.4
also rather throws the book at the NFW profile relations – it probably isn't
necessary to include equation 22."]}
As mentioned above, JAM modelling allows to incorporate an invisible matter component in addition to the luminous matter in the form of an MGE. In Section \ref{sec:results_JAM_NFW} we will include a spherical Navarro-Frenk-White (NFW) DM halo \citep{1996ApJ...462..563N} in the dynamical model. The classical NFW profile has the form
\begin{equation}
\rho_\text{NFW}(r) \propto r^{-1} \left( r+r_s \right)^{-2} \label{eq:NFWprofile}
\end{equation}
and two free parameters, the scale length $r_s$ and a parameter describing the total mass of the halo. We will use $v_\text{200}$, which is the circular velocity at the radius $r_\text{200}$ within which the mean density of the halo is 200 times the cosmological critical density $\rho_\text{crit} \equiv (2H^2)/(8\pi G)$, i.e.,
\begin{eqnarray*}
M_\text{200} &=& M(<r_{200})\\
\frac{M_{200}}{ \frac 43 \pi r_{200}^3} &=& 200 \rho_\text{crit}(z=0) \\
v_\text{200} &=& \sqrt{\frac{GM_{200}}{r_\text{200}}}
\end{eqnarray*}
with $\rho_\text{crit}(z=0)=1.43 \cdot 10^{-7} M_\odot / \text{pc}^3$ in the WMAP5 cosmology by \citet{WMAP5cosm}. How much the mass is concentrated in the center of the NFW halo is given by the concentration parameter
\begin{equation}
c_{200}\equiv r_{200} / r_s. \label{eq:NFW_c}
\end{equation} 
There is a close relation between the concentration and halo mass in simulations \citep{1996ApJ...462..563N}. \citet{Maccio08} found this relation for the WMAP5 cosmology to be
\begin{equation}
\langle \log c_{200} \rangle (M_{200}) = 0.830 - 0.098 \log \left(h \frac{M_{200}}{10^{12} M_\odot} \right) \label{eq:Maccio08}
\end{equation}
(their equation 10), with a Gaussian scatter of $\sigma_{\log c_{200}} = 0.105$ (their Table A2). 
%--------------------------------------------

@@

Figure \ref{fig:JAM_modelG} shows that the best fit model nicely reproduces the central dip in the $v_\text{rms}$ curve, even though it has difficulties fitting the drop around $R' \sim 4''$. The latter is because we only allowed the $\Upsilon_\text{I,tot}(R')$ to rise. A slight drop could be expected when the reddish bulge turns into the bluish disk and the contribution of the stellar component becomes less due to a lower $\Upsilon_\text{I,*}$ for younger and bluer populations.

In Figure \ref{fig:enclMass_modelG} we overplot the enclosed mass profile with the Einstein mass $M_\text{ein} = (7.77 \pm 0.33) \cdot 10^{10} M_\odot$ at the Einstein radius found from lensing in Table \ref{tab:bestfitlensmodel}. The agreement between the Einstein mass and the independently found $M(<R_\text{ein}) = 7.49 \cdot 10^{10} M_\odot$ from dynamical modelling is striking.
\Wilma{[TO DO: referee says: "From a scientific perspective, in section 4.3.3, the authors comment that the
varying M/L JAM model is unable to reproduce the dynamics when M/L is forced to be
always rising as a function of radius, but it would be interesting to see if the
model could be improved by removing this constraint. Perhaps the authors could add
such a model to this section, or else comment on why they do not attempt it." --> Add sentence that with 5 fit parameters and 6 data points within 5 arcsec and no additional information it is likely that we perform overfitting: The resulting best fit profile is wiggly and hard to interpret physically.]}
%--------------------------------------------

@@

@@

We restrict the $v_\text{rms}$ fit to a region $R' \lesssim 3.5''$, approximately within the effective half-light radius $R_\text{eff} = 2.6''$. We also include the Einstein mass $M(<R_\text{ein})$ with a 10\% error as an additional constraint in the fit.
%--------------------------------------------

@@

While the dispersion $\sigma$ in the center fits by construction quite good, the predicted dispersion is much larger than the data. We would expect the disk to be rotationally supported and therefore have a low velocity dispersion. Especially dispersions as high as $\sim 200~\text{km s}^{-1}$ are more likely to be observed in the pressure supported bulges of galaxies. There might be something unexpected with the $\sigma$ measurements around $\sim 5''$, but at large radii the best fit model NFW halo is simply too massive.
%--------------------------------------------

@@

Overall, we were not able to find a consistent explanation for J1331's central $v_\text{rms}$ dip. It cannot be due to tangential velocity anisotropy in the center alone (see Section \ref{sec:results_JAM_SB_MfL}, Figure \ref{fig:JAM_modelA2}). It also cannot be explained by a strong contribution of DM inside the bulge together with a moderate tagential velocity anisotropy, because the corresponding DM halos would still be too massive to fit the data in the outer regions (see Section \ref{sec:results_JAM_NFW}, Figure \ref{fig:modelB4_vrms}).

@@

>>>>

%--------------------------------------------
\section{Introduction} \label{sec:intro}
%--------------------------------------------

@@

%--------------------------------------------
As mentioned above, JAM modelling allows to incorporate an invisible matter component in addition to the luminous matter in the form of an MGE. In Section \ref{sec:results_JAM_NFW} we will include a spherical Navarro-Frenk-White (NFW) DM halo \citep{1996ApJ...462..563N} in the dynamical model. The classical NFW profile has the form
\begin{equation}
\rho_\text{NFW}(r) \propto r^{-1} \left( r+r_s \right)^{-2} \label{eq:NFWprofile}
\end{equation}
and two free parameters, the scale length $r_s$ and a parameter describing the total mass of the halo. We will use $v_\text{200}\NEW{\equiv \sqrt{GM_{200}/r_{200}}}$, which is the circular velocity at the radius $r_\text{200}$ within which the mean density of the halo is 200 times the cosmological critical density $\rho_\text{crit} \equiv (2H^2)/(8\pi G)$, \OLD{i.e.,}\NEW{ and where $M_{200} \equiv M(<r_{200}) \stackrel{!}{=} \frac 43 \pi r_{200}^3 \cdot \rho_\text{crit}(z=0)$.}
\OLD{\begin{eqnarray*}
M_\text{200} &=& M(<r_{200})\\
\frac{M_{200}}{ \frac 43 \pi r_{200}^3} &=& 200 \rho_\text{crit}(z=0) \\
v_\text{200} &=& \sqrt{\frac{GM_{200}}{r_\text{200}}}
\end{eqnarray*}}
with $\rho_\text{crit}(z=0)=1.43 \cdot 10^{-7} M_\odot / \text{pc}^3$ in the WMAP5 cosmology by \citet{WMAP5cosm}. \OLD{How much the mass is concentrated in the center of the NFW halo is given by the concentration parameter
\begin{equation}
c_{200}\equiv r_{200} / r_s. \label{eq:NFW_c}
\end{equation}}
\NEW{The mass concentration of the NFW halo is quantified by $c_{200}\equiv r_{200} / r_s$.}
There is a close relation between the concentration and halo mass in simulations \citep{1996ApJ...462..563N}. \citet{Maccio08} found this relation for the WMAP5 cosmology to be
\begin{equation}
\langle \log c_{200} \rangle (M_{200}) = 0.830 - 0.098 \log \left(h \frac{M_{200}}{10^{12} M_\odot} \right) \label{eq:Maccio08}
\end{equation}
(their equation 10), with a Gaussian scatter of $\sigma_{\log c_{200}} = 0.105$ (their Table A2). \NEW{In Section \Wilma{[TO DO ???]} we will use this relation as prior information to guide the modelling towards realistic NFW halo shapes.}
%--------------------------------------------

@@

Figure \ref{fig:JAM_modelG} shows that the best fit model nicely reproduces the central dip in the $v_\text{rms}$ curve, even though it has difficulties fitting the drop around $R' \sim 4''$. The latter is because we only allowed the $\Upsilon_\text{I,tot}(R')$ to rise. A slight drop could be expected when the reddish bulge turns into the bluish disk and the contribution of the stellar component becomes less due to a lower $\Upsilon_\text{I,*}$ for younger and bluer populations. \NEW{Corresponding fitting attempts with the matter model at hand were however unsuccessful: Leaving all five $\Upsilon_{\text{I,tot,}i}$ free lead to over-fitting in the central regions of J1331, while not allowing for sufficient flexibility further out at $R'\gtrsim4''$ (because four of the five $\sigma_i$ in the light MGE in Table \ref{tab:MGEF814W} are $< 1.2''$).}

\NEW{For the inner regions the model with rising $\Upsilon(R')$ seems to be a very good model:} In Figure \ref{fig:enclMass_modelG} we overplot the enclosed mass profile with the Einstein mass $M_\text{ein} = (7.77 \pm 0.33) \cdot 10^{10} M_\odot$ at the Einstein radius found from lensing in Table \ref{tab:bestfitlensmodel}. The agreement between the Einstein mass and the independently found $M(<R_\text{ein}) = 7.49 \cdot 10^{10} M_\odot$ from dynamical modelling is striking.
%--------------------------------------------

@@

\NEW{As we are particularly interested in the disentanglement of DM and stellar matter in the inner regions (see Section \ref{sec:intro}), we restrict the $v_\text{rms}$ fit to a region $R'\lesssim3.5''$, approximately within the effective half-light radius $R_\text{eff}=2.6''$. The constraints on a DM halo just from this data might be weak. To guide the fit towards a realistic halo, we therefore impose several priors on NFW halo.}

@@

\OLD{We restrict the $v_\text{rms}$ fit to a region $R' \lesssim 3.5''$, approximately within the effective half-light radius $R_\text{eff} = 2.6''$. }We also include the Einstein mass $M(<R_\text{ein})$ with a 10\% error as an additional constraint in the fit.
%--------------------------------------------

@@

While the dispersion $\sigma$ in the center fits by construction quite good, the predicted dispersion is much larger than the data. We would expect the disk to be rotationally supported and therefore have a low velocity dispersion. Especially dispersions as high as $\sim 200~\text{km s}^{-1}$ are more likely to be observed in the pressure supported bulges of galaxies. There might be something unexpected with the $\sigma$ measurements around $\sim 5''$, but at large radii the best fit model NFW halo is simply too massive.

\subsubsection{\NEW{Discussion}}

\NEW{A fit to all nine available $v_\text{rms}$ data points forces the DM halo to be less massive. But with decreasing DM contribution in the inner regions the model approaches the mass-follows-light model in the center, that we already ruled out in Section \ref{sec:results_JAM_SB_MfL}. Allowing for free $\Upsilon_{*}(R')$, $\beta_z(R')$ and/or non-NFW $\rho_\text{halo}(r)$ profiles would over-fit the data again. Without further priors on the stellar $\Upsilon_{*}(R')$ and/or more data points, we can make only the following statement: A model with constant $\Upsilon_*$, moderately tangential \Wilma{[TO DO: Check]} $\beta_z$ and a spherical and massive DM component with high concentration can explain the central kinematics of J1331 very well, we have however ruled out that in this case the DM halo will follow a NFW halo at both small and large radii.}
%--------------------------------------------

@@

Overall, we were not able to find a consistent explanation for J1331's central $v_\text{rms}$ dip. It cannot be due to tangential velocity anisotropy in the center alone (see Section \ref{sec:results_JAM_SB_MfL}, Figure \ref{fig:JAM_modelA2}). It also cannot be explained by a strong contribution of DM inside the bulge together with a moderate tagential velocity anisotropy, because the corresponding \NEW{NFW} DM halo\OLD{s} would still be too massive to fit the data in the outer regions (see Section \ref{sec:results_JAM_NFW}, Figure \ref{fig:modelB4_vrms}).

@@

\subsection{\NEW{Speculations on J1331's matter distribution}}

\NEW{The kinematics of the merger remnant J1331 are peculiar and hard to explain with standard modelling techniques. We think that in the end the truth will be somewhere in between the results of the different modelling attempts.}

\NEW{The central $v_\text{rms}$ is most likely a combination of the three different aspects that we investigated separately: We expect a bit of tangential \Wilma{???} velocity anisotropy (Sections \ref{sec:results_JAM_SB_MfL} and \ref{sec:results_JAM_NFW}), a moderate contribution of a roundish DM halo (Sections 4.2.3 \Wilma{???} and 4.4.2) and different stellar populations inside the bulge due to the merger (Section 5.3) with different $\Upsilon_{*}$ but overall bottom-heavy IMF (Section 5.1), which add to an overall rising $\Upsilon_\text{tot}(R')$ profile.}

\NEW{The drop in $v_\text{rms}$ around $R'\sim4''$ is most likely due to a sharp drop in $\Upsilon_{*}(R')$ at the transition from red bulge to blue disk (compare bulge size in Figures \ref{fig:F450W} and \ref{fig:F814W}). Unfortunately the spiral arms make the disk of the galaxy difficult to model. At these larger radii a possible mis-alignment of long-slit and Galaxy major axis would also cause a stronger underestimation of the true $v_\text{rot}$ and would add to the drop in $v_\text{rms}$.}

\NEW{Around $R'\sim7''$ the rotation curve and dispersion show some wiggles which lead to a spontaneous rise of the $v_\text{rms}$. As the spiral arms cross the major axis around this radius, we suspect they cause this disturbance of the axisymmetric kinematics. The DM halo should also start to dominate the kinematics at larger radii, whose total mass we expect more of the order of what \citep{Dutton paper} \Wilma{???} found---unless J1331 has a strong kinematic warp in its disk, which cannot be detected in long-slit kinematics only.}

\NEW{All of these speculations should be easily testable with 2D kinematics.} \Wilma{[TO DO: Proper referencing.]}
